### 2017-10-12 thesis/discussion/discussion.tex
Final revision completed

The fact that both CR and FM obesity-associated genetic signatures showed no sign of significant association with the patient \gls{bmi} in majority of the cancer data sets raised a question of whether these signatures were actually related to obesity, or a different clinical variable within the data set.
This question led to the identification of various versions of obesity-associated genetic signatures in \cref{sub:identification_of_obesity_associated_genetic_signatures}.
The CR data set was controlled for all the clinical variables (sex, age, ethnicity, menopause status, tumour grade, \gls{er}/\gls{her2}/\gls{pr} statuses and \gls{ln} status) to generate some of these signatures.
All of the metagenes created from these novel obesity-associated genetic signatures were consistent with the gene expression patterns, but none of the metagenes showed significant association with the patient \gls{bmi} or \gls{bmi} status in the data set other than the CR data set, with the exception of the overweight group in the \gls{blca} data set.

It was possible that all of these results were due to the sample quality (and hence due to the quality of the data), or perhaps many of the samples in these data sets had breast cancers due to a cause unrelated to their obesity status.
As many epidemiological studies have pointed out, obesity is a strong risk factor for cancer (covered in \cref{sec:obesity_and_cancer}), but this does not necessarily mean that all cancer in obese patients develop cancer as a result of obesity.
As briefly mentioned in \cref{ssub:environmental_factors}, there are many other environmental factors that can cause cancer and patients in these studies may have been exposed to these factors, but not recorded or measured the exposure in the clinical data for adjustment in the analysis.

In other words, the cancer identified in obese patients may have been caused by another, completely different biological mechanism.
Thus the tumour samples taken from obese and overweight patients will be a mixture of samples that are driven by obesity-related mechanisms and mechanisms that are not related to obesity.
These different subgroups of tumours (obesity-driven or not) cannot be easily identified and isolated.
This may affect the detection of the truly obesity-associated genes, as some of the tumour samples from the patients will have nothing to do with obesity and cancer, and hence misclassified into wrong groups.

@@

With that said, there is a possibility that the genotypes of the patients that are overweight in the \gls{blca} data set are similar to those patients that are obese in the CR data set.
In fact, \citet{Damrauer2014} have identified ``basal-like'' and ``luminal''  subtypes in bladder cancer that resembled the corresponding subtypes in breast cancer, and suggested that bladder  and breast cancers may have common physiological properties.
Since all of the obesity-associated genetic signatures were derived from breast cancer data sets, it made sense that these signatures were significant in the \gls{blca} data set, as bladder cancer and breast cancer could have been biologically similar to one another.
However, the degree of similarity between these two cancer types are not yet known and further analyses are required for a definitive explanation of the results.

@@

From all of these results so far, one may question how robust or appropriate the use of \gls{svd} and transformation matrices are to generate metagenes in other data sets.
It appeared as though the CR data-derived obesity-associated genetic signatures correlated significantly in the CR data, but as soon as these metagenes were generated in the other data sets via the application of transformation matrices, the correlation became insignificant.
One possible reason for this was that the obesity metagenes were not actually associated to the obesity phenotype in the other data sets.
Another reason could be that the methodology used was not appropriate for investigating the association between the phenotype and its associated phenotype.

@@

The fact that there were so many false positives in the \glspl{deg} from the \gls{icgc} data sets indicated that the use of the patient \gls{bmi} and \gls{bmi} status as the ``treatment'' conditions (in other words independent variable) for gene expression analysis was prone to \Glspl{type1}.
This meant that there were very little evidence of true obesity-associated genetic signatures in any of the data sets that have been explored in this project, or perhaps the patient \gls{bmi} and \gls{bmi} status was not enough to pinpoint the underlying biological relationship between obesity and cancer (see \cref{sub:discussion_definition_of_obesity} as well).

Evidence of large number of false positives was also seen in \cref{sub:identification_of_obesity_associated_genetic_signatures}, where there were many \glspl{deg} identified from the gene expression analyses when obesity-associated genetic signatures were sought for in the CR data set.
This suggested that there was a significant number of genes that were not related to obesity (false positives) in the genetic signatures.
Furthermore, the fact that about a third of the genes in any one of the obesity signatures were ``unique'' to each of the individual obesity signatures supported the apparent abundance of false positives in the signatures (\cref{fig:venn1,fig:venn2}).
From these results, it was likely that the obesity signatures that resulted from the CR data set had many genes that were false positives, perhaps due to the poor quality of the data set (\cref{sub:quality_of_the_data}).

@@

In \cref{cha:obesity_associated_genetic_signature_and_pathway_signatures}, analyses were carried out to determine what the genetic pattern the obesity metagenes from \cref{cha:obesity_genetic_signatures_and_cancer} were associating with.
The pathway-associated genetic signatures from the \citet{Gatza2010a} study were used to establish the biological connection with the obesity-associated genetic signatures.
In order to make the obesity- and pathway-associated genetic signatures comparable with one another, the orientations of the metagenes were examined so that all of the metagenes were in the ``correct'' orientation (see \cref{sec:pathway_associated_genetic_signatures_from_gatza2010a_study}).

@@

To further investigate whether there was any evidence of a pathway signature being related to the obesity-associated genetic signatures, linear models were created in the \gls{nzbc} data set with the pathway metagene scores.
First, linear models were created with the patient \gls{bmi}, \gls{bmi} status and a selection of the most ``consistent'' pathway metagene scores (based on the consistency across different data sets).
From these linear models, \gls{pr} pathway metagene scores showed significant contribution to the model in predicting the obesity-associated metagene scores in all of the linear models, with the exception of the FM obesity metagene (where Myc was the significant variable; \cref{sec:summary_of_the_linear_models_in_nzbc_data}).

@@

The linear models that were created from the ``consistent'' pathway signatures in the \gls{nzbc} data set showed significant association of the \gls{pr} pathway with the obesity metagenes.
But was this due to the true biological relationship between the \gls{pr} pathway and the obesity metagenes, or only because the models were created in the \gls{nzbc} data set?
In fact, the proportion of the patients that were \gls{pr}$^+$ was greater in the \gls{nzbc} data set (62 \gls{pr}$^+$ and 37 \gls{pr}$^-$ patients) compared with the CR data set (48 \gls{pr}$^+$ and 50 \gls{pr}$^-$ patients), which implied that the patients in the \gls{nzbc} data set were more likely to show association with the \gls{pr} pathway metagene.

@@

In the step wise linear models, many variables were identified to be significantly associated with the obesity metagenes.
This suggested that the pathway-associated genetic signatures that were significant in these models may be biologically related to the obesity phenotype, or perhaps the obesity-associated genetic signatures were detecting the signals from these pathways rather than the signals from the obesity phenotype.
In fact, many of the variables included in the linear models were pathway metagenes that negatively correlated with the obesity metagenes in \cref{fig:gatza_allmeta} (\cref{sub:comparison_of_the_obesity_and_pathway_associated_signatures}).
Since all of the obesity metagenes failed to show significant association with patient \gls{bmi}, it may be possible that these pathways (Akt, \gls{er}, \gls{egfr}, \gls{pr}, p53, Src and \gls{tgfb}) are actually what was being detected by the ``obesity'' metagenes.

>>>>

The fact that neither CR nor FM obesity-associated genetic signatures showed any sign of significant association with the patient \gls{bmi} in the majority of the cancer data sets raised a question of whether these signatures were actually related to obesity, or to a different clinical variable within the data set.
This question led to the identification of various versions of obesity-associated genetic signatures in \cref{sub:identification_of_obesity_associated_genetic_signatures}.
The CR data set was controlled for all the clinical variables (sex, age, ethnicity, menopause status, tumour grade, \gls{er}/\gls{her2}/\gls{pr} statuses and \gls{ln} status) to generate some of these signatures.
All of the metagenes created from these novel obesity-associated genetic signatures were consistent with the gene expression patterns, but none of the metagenes showed significant association with the patient \gls{bmi} or \gls{bmi} status in the data set other than the CR data set, with the exception of the overweight group in the \gls{blca} data set.

It was possible that all of these results were due to the sample quality (and hence due to the quality of the data), or perhaps many of the samples in these data sets had breast cancers due to a cause unrelated to their obesity status.
As many epidemiological studies have pointed out, obesity is a strong risk factor for cancer (covered in \cref{sec:obesity_and_cancer}), but this does not necessarily mean that all cancer in obese patients develop cancer as a result of obesity.
As briefly mentioned in \cref{ssub:environmental_factors}, there are many other environmental factors that can cause cancer, and patients in these studies may have been exposed to these factors, but this exposure information is not recorded in the clinical data, so cannot be accounted for in the analysis.

In other words, the cancer identified in obese patients may have been caused by another, completely different biological mechanism.
Thus the tumour samples taken from obese and overweight patients will be a mixture of samples that are driven by obesity-related mechanisms and mechanisms that are not related to obesity.
These different subgroups of tumours (obesity-driven or not) cannot be easily identified and isolated.
This may affect the detection of the truly obesity-associated genes, as some of the tumour samples from the patients will have nothing to do with obesity and cancer, and hence will be misclassified into the wrong groups.

@@

With that said, there is a possibility that the genotypes of the patients that are overweight in the \gls{blca} data set are similar to those patients that are obese in the CR data set.
In fact, \citet{Damrauer2014} have identified ``basal-like'' and ``luminal''  subtypes in bladder cancer that resembled the corresponding subtypes in breast cancer, and suggested that bladder  and breast cancers may have common physiological properties.
Since all of the obesity-associated genetic signatures were derived from breast cancer data sets, it makes some sense that these signatures were significant in the \gls{blca} data set, as there is evidence that bladder cancer and breast cancer are biologically similar to one another \citep{Damrauer2014}.
However, the degree of similarity between these two cancer types are not yet known and further analyses are required for a definitive explanation of the results.

@@

From all of these results so far, one may question how robust or appropriate the use of \gls{svd} and transformation matrices are to generate metagenes in other data sets.
The CR data-derived obesity-associated genetic signatures correlated significantly in the CR data, but as soon as these metagenes were generated in the other data sets via the application of transformation matrices, the correlation became non-significant.
One possible reason for this was that the obesity metagenes were not actually associated to the obesity phenotype in the other data sets.
Another reason could be that the methodology used was not appropriate for investigating the association between the phenotype and its associated phenotype.

@@

The fact that there were so many false positives in the \glspl{deg} from the \gls{icgc} data sets indicated that the use of the patient \gls{bmi} and \gls{bmi} status as the ``treatment'' conditions (in other words independent variable) for gene expression analysis was prone to \Glspl{type1}.
This meant that there was very little evidence of true obesity-associated genetic signatures in any of the data sets that were  explored in this project, or perhaps the patient \gls{bmi} and \gls{bmi} status was not enough to pinpoint the underlying biological relationship between obesity and cancer (also see \cref{sub:discussion_definition_of_obesity}).

Evidence of a large number of false positives was also seen in \cref{sub:identification_of_obesity_associated_genetic_signatures}, where there were many \glspl{deg} identified from the gene expression analyses when obesity-associated genetic signatures were sought for in the CR data set.
This suggested that there was a significant number of genes that were not related to obesity (false positives) in the genetic signatures.
Furthermore, the fact that about a third of the genes in any one of the obesity signatures were ``unique'' to each of the individual obesity signatures supported the apparent abundance of false positives in the signatures (\cref{fig:venn1,fig:venn2}).
From these results, it was likely that the obesity signatures that resulted from the CR data set had many genes that were false positives, perhaps due to the poor quality of the data set (\cref{sub:quality_of_the_data}).

@@

In \cref{cha:obesity_associated_genetic_signature_and_pathway_signatures}, analyses were carried out to determine what the genetic pattern the obesity metagenes from \cref{cha:obesity_genetic_signatures_and_cancer} were associating with.
The pathway-associated genetic signatures from the \citet{Gatza2010a} study were used to establish the biological connection with the obesity-associated genetic signatures.
In order to make the obesity- and pathway-associated genetic signatures comparable with one another, the orientations of the metagenes were examined so that all of the metagenes were in a consistent orientation (see \cref{sec:pathway_associated_genetic_signatures_from_gatza2010a_study}).

@@

To further investigate whether there was any evidence of a pathway signature being related to the obesity-associated genetic signatures, linear models were created in the \gls{nzbc} data set with the pathway metagene scores.
First, linear models were created with the patient \gls{bmi}, \gls{bmi} status and a selection of the most ``consistent'' pathway metagene scores (based on the consistency across different data sets).
From these linear models, \gls{pr} pathway metagene scores showed a significant contribution to the model in predicting the obesity-associated metagene scores in all of the linear models, with the exception of the FM obesity metagene (where Myc was the significant variable; \cref{sec:summary_of_the_linear_models_in_nzbc_data}).

@@

The linear models that were created from the ``consistent'' pathway signatures in the \gls{nzbc} data set showed significant association of the \gls{pr} pathway with the obesity metagenes.
But was this due to the true biological relationship between the \gls{pr} pathway and the obesity metagenes, or only because the models were created in the \gls{nzbc} data set?
In fact, the proportion of patients that were \gls{pr}$^+$ was greater in the \gls{nzbc} data set (62 \gls{pr}$^+$ and 37 \gls{pr}$^-$ patients) compared with the CR data set (48 \gls{pr}$^+$ and 50 \gls{pr}$^-$ patients), which implied that the patients in the \gls{nzbc} data set were more likely to show association with the \gls{pr} pathway metagene.

@@

In the step wise linear models, many variables were identified to be significantly associated with the obesity metagenes.
This suggested that the pathway-associated genetic signatures that were significant in these models may be biologically related to the obesity phenotype, or perhaps the obesity-associated genetic signatures were detecting signals from these pathways rather than the signals from the obesity phenotype.
In fact, many of the variables included in the linear models were pathway metagenes that negatively correlated with the obesity metagenes in \cref{fig:gatza_allmeta} (\cref{sub:comparison_of_the_obesity_and_pathway_associated_signatures}).
Since all of the obesity metagenes failed to show significant association with patient \gls{bmi}, it may be possible that these pathways (Akt, \gls{er}, \gls{egfr}, \gls{pr}, p53, Src and \gls{tgfb}) are actually what was being detected by the ``obesity'' metagenes.

### 2017-10-12 thesis/intro/intro.tex
Final revision completed

\noindent
Approximately 1 in every 14 people in the world were classed as obese in 2014; therefore, 1 in every 14 people around the world were also at severe health risks associated with obesity.
The highest prevalence of overweight and obesity was in the American regions, where 61\% and 27\% of the population were overweight and obese respectively \citep{WHO2014}.
It was also noted in the report that women were more likely to be obese than men, and that the prevalence of overweight and obesity increased as the income level of the countries increased \citep{WHO2014}.

@@

In addition to the mechanisms described above, there is a growing body of evidence where some environmental factors may affect the transcriptional regulation of a set of genes in certain cancer type.
A relatively recent review by \citet{Perez-Solis2016} describes potential mechanism in which obesity, alcoholism and smoking (all a lifestyle factor related to obesity) may regulate gene expression in breast cancer cells.
By affecting the so-called ``Master'' transcription factors, environmental factors are able to affect the expression of the genes that are downstream to the Master transcription factor, and as a result, lead to the development of cancerous cells \citep{Perez-Solis2016}.
It is not surprising that different environmental factors affect different Master transcriptional factor, and therefore have different biological mechanism of gene regulation.

It is also worth noting here that multiple  environmental factors may be exposed to an individual at any given point of time, and perhaps a combination of multiple environmental factors may have more significant effect on the individual.
There are many other environmental factors associated with cancers that have not been mentioned here.
Nevertheless, it is evident that limiting the exposure to detrimental environmental factors is crucial for the prevention of many cancers.

@@

Adipose tissue has long known to have been a group of cells to store excess fats, but it has recently become apparent that adipose tissue is also a metabolically active endocrine organ \citep{Roberts2010}.
Increased adiposity is known to influence the bioavailability of endogenous sex hormones, including estrogens, androgens and progesterone \citep{Calle2004}.
\Glspl{adipocyte} synthesise enzymes, such as aromatase, that convert precursor molecules into different forms of androgens and estrogens \citep{Calle2004}.
Many studies have confirmed the mitogenic effect of sex hormones on certain cell types, mainly breast and endometrial cell types \citep{Roberts2010}.
In addition to this, there has also been evidence of serum from obese patient influencing the gene expression of sex hormone receptor (estrogen receptor $\beta$) in a breast cancer cell line, suggesting there may be a gene regulatory effect caused by the obese phenotype \citep{Bowers2015}.
Through the effects of dysregulated levels of hormones and hormone receptors, obesity may have significant impact on the tumour biology through changes in gene expression level.

@@

Major concern with this study was the quality of the \acrshort{rna} that was collected from the samples.
Ensuring that all of the samples had more than 70\% of tumour cells meant that the \acrshort{rna} was likely from the tumour cells.
However, this did not control for the quality of the \acrshort{rna} that has been extracted, as some samples may have had significantly less tumour volume compared to another, but still have $\geq$ 70\% tumour cells in both samples.
Another thing to note from this paper was the predictive ability of their obesity signatures in other data sets.
\citet{Creighton2012} used six independent breast cancer  data sets to test whether their signature was able to replicate their findings.
Although they were able to replicate their results in one of the data sets, they were not able to find any significant association in the other data sets.

>>>>

\noindent
Approximately 1 in every 14 people in the world were classed as obese in 2014; therefore, 1 in every 14 people around the world were also at risk of the severe health consequences associated with obesity.
The highest prevalence of overweight and obesity was in the American regions, where 61\% and 27\% of the population were overweight and obese respectively \citep{WHO2014}.
It was also noted in the report that women were more likely to be obese than men, and that the prevalence of overweight and obesity increased as the income level of the countries increased \citep{WHO2014}.

@@

In addition to the mechanisms described above, there is a growing body of evidence where some environmental factors may affect the transcriptional regulation of a set of genes in certain cancer type.
A relatively recent review by \citet{Perez-Solis2016} describes potential mechanisms in which obesity, alcoholism and smoking (all lifestyle factors related to obesity) may regulate gene expression in breast cancer cells.
By affecting the so-called ``Master'' transcription factors, environmental factors are able to affect the expression of the genes that are downstream to the Master transcription factor, and as a result, lead to the development of cancerous cells \citep{Perez-Solis2016}.
It is not surprising that different environmental factors affect different Master transcriptional factors, and therefore have different biological mechanisms of gene regulation.

It is also worth noting here that an individual may be exposed to  multiple  environmental factors at any given point of time, and perhaps a combination of multiple environmental factors may have more significant effect on the individual.
There are many other environmental factors associated with cancers that have not been mentioned here.
Nevertheless, it is evident that limiting the exposure to detrimental environmental factors is crucial for the prevention of many cancers.

@@

Adipose tissue has long known to have been a group of cells to store excess fats, but it has recently become apparent that adipose tissue is also a metabolically active endocrine organ \citep{Roberts2010}.
Increased adiposity is known to influence the bioavailability of endogenous sex hormones, including estrogens, androgens and progesterone \citep{Calle2004}.
\Glspl{adipocyte} synthesise enzymes, such as aromatase, that convert precursor molecules into different forms of androgens and estrogens \citep{Calle2004}.
Many studies have confirmed the mitogenic effect of sex hormones on certain cell types, mainly breast and endometrial cell types \citep{Roberts2010}.
In addition to this, there has also been evidence of serum from obese patients influencing the gene expression of sex hormone receptor (estrogen receptor $\beta$) in a breast cancer cell line, suggesting there may be a gene regulatory effect caused by the obese phenotype \citep{Bowers2015}.
Through the effects of dysregulated levels of hormones and hormone receptors, obesity may have significant impact on the tumour biology through changes in gene expression level.

@@

One major concern with this study was the quality of the \acrshort{rna} that was collected from the samples.
Ensuring that all of the samples had more than 70\% of tumour cells meant that the \acrshort{rna} was likely from the tumour cells.
However, this did not control for the quality of the \acrshort{rna} that was extracted, as some samples may have had significantly less tumour volume compared to another, but still have $\geq$ 70\% tumour cells in both samples.
Another thing to note from this paper was the predictive ability of their obesity signatures in other data sets.
\citet{Creighton2012} used six independent breast cancer  data sets to test whether their signature was able to replicate their findings.
Although they were able to replicate their results in one of the data sets, they were not able to find any significant association in the other data sets.

### 2017-10-12 thesis/methods/methods.tex
Final revision completed

The raw Affymetrix HGU\-133A microarray gene expression data files from the \citet{Creighton2012} study were downloaded from the \gls{geo} database (\gls{geo} accession ID: GSE24185).
Clinical data of the samples (age, ethnicity, tumour grade, menopause status, \gls{bmi}, \gls{er} status, \gls{pr} status, \gls{her2} status, and \gls{ln} status) were obtained from the Supplementary Table 1 from the \citet{Creighton2012} study.
799 obesity-associated gene probes identified in the \citet{Creighton2012} study were obtained from the Supplementary Data File 1 from the \citet{Creighton2012} study.

@@

The raw Affymetrix HGU\-133A microarray gene expression data files from the  \citet{Fuentes-Mattei2014} study were downloaded from the \gls{geo} database (\gls{geo} accession ID: GSE\-20194).
Clinical data for the samples (age, ethnicity, tumour grade, \gls{er}/\gls{pr}/\gls{her2} statuses, and treatments used) in this study was also downloaded from the \gls{geo} database (same \gls{geo} accession ID).
Patient height, weight and \gls{bmi} information were not available in this data set.
130 obesity-associated gene probes identified by \citet{Fuentes-Mattei2014} were taken from the Supplementary Table 3 from their paper.

@@

The raw microarray gene expression data files from the \citet{Gatza2010a} study were downloaded from the \gls{geo} database (\gls{geo} accession ID: GSE1456, GSE\-1561, GSE2034, GSE3494, GSE4922 and GSE6596).
Only the Affymetrix HGU\-133A microarray samples were included in this project, as the other microarray data from the \citet{Creighton2012}, \citet{Fuentes-Mattei2014} and \citet{Print2016} studies were analysed using the Affymetrix HGU-133A platform.
Clinical data for the samples in \citet{Gatza2010a} study was not available, as these samples were a combination of many different datasets.
The Akt, \gls{bcat}, E2F1, \gls{egfr}, \gls{er}, \gls{her2}, \gls{ifna}, \gls{ifny}, Myc, p53, p63, \gls{pi3k}, \gls{pr}, Ras, \gls{stat3}, Src, \gls{tgfb}, and \gls{tnfa} pathway-associated genetic signatures were used by Gatza \textit{et al.}, and the gene probes of all the pathway-associated genetic signatures were obtained from the Supplemental Table 10 from their paper.

@@

The raw microarray data were normalised using both \gls{rma} and \gls{mas} methods, each done separately on a copy of the data.
The reason why the \gls{mas} method was used as well as the \gls{rma} method was because the \gls{mas}-normalisation method was used in the \citet{Gatza2010a} study.
Though the normalisation methods were not specified in the \citet{Creighton2012} and \citet{Fuentes-Mattei2014} studies, to allow for data comparison between different datasets and accurate result validation of some of the studies, both \gls{rma} and \gls{mas} methods were considered for normalising the microarray data.

@@

In this project, two types of metagenes were created: obesity-associated and pathway-associated metagenes.
For obesity-associated metagenes, the direction of the metagenes were checked with the patient obesity status and \gls{bmi} value from the data in which the metagene was derived from.
When the data set did not have \gls{bmi} or \gls{bmi} status information for the patients (for example the Gatza \textit{et al.} data set), a set of genes that were common in all obesity-associated genetic signatures were taken and the expression of these common genes were compared with the metagene scores.
The genes that are common to all of the obesity-associated genetic signatures will be related to the metagenes produced by these signatures, and therefore the expressions of these genes should be in line with the metagene scores.

Checking the direction of the pathway metagenes from \citet{Gatza2010a} study is not as simple as the obesity-associated metagenes, as the experimental design that identify which samples were treated to generate the pathway signature was not available.
Due to this, it was not possible to validate whether the generated pathway metagene was in the ``correct'' direction.
As an alternative, the pathway metagenes were visually compared with the gene that represent the pathway using a heatmap, as the expression of this gene would likely be affected the most by the pathway signature.
For example, the \gls{er} pathway metagene was compared with the \textit{ESR1} gene expression, and if the metagene was not in the same direction as the \textit{ESR1} gene expression, the direction of the metagene was corrected (\cref{fig:methods/meta_dir}).
\cref{tab:metagene_direction} summarises the genes used to check the direction of each of the pathway metagenes.

@@

Since the \gls{nzbc} data was the only data set (other than the Creighton \textit{et al.} data set) that had the patient \gls{bmi} information, all of the linear models were constructed in the \gls{nzbc} data set.
Eleven linear models were created in the \gls{nzbc} to predict the obesity-associated metagene from the \citet{Creighton2012} study.
First seven models created were combinations of the sample \gls{bmi}, \gls{bmi} status and a selection of pathway-associated metagene scores.
The selected pathways used in the linear model construction were \gls{bcat}, \gls{er}, \gls{ifna}, \gls{ifny}, Myc and \gls{pr}.
In deciding on the pathways to be used for the construction of the linear model, the correlation of the \gls{svd}- and transformation matrix-derived pathway metagene scores in different data sets were examined.

>>>>

The raw Affymetrix HGU\-133A microarray gene expression data files from the \citet{Creighton2012} study were downloaded from the \gls{geo} database (\gls{geo} accession ID: GSE24185).
Clinical data of the samples (age, ethnicity, tumour grade, menopause status, \gls{bmi}, \gls{er} status, \gls{pr} status, \gls{her2} status, and \gls{ln} status) were obtained from Supplementary Table 1 of the \citet{Creighton2012} study.
799 obesity-associated gene probes identified in the \citet{Creighton2012} study were obtained from Supplementary Data File 1 from the \citet{Creighton2012} study.

@@

The raw Affymetrix HGU\-133A microarray gene expression data files from the  \citet{Fuentes-Mattei2014} study were downloaded from the \gls{geo} database (\gls{geo} accession ID: GSE\-20194).
Clinical data for the samples (age, ethnicity, tumour grade, \gls{er}/\gls{pr}/\gls{her2} statuses, and treatments used) in this study was also downloaded from the \gls{geo} database (same \gls{geo} accession ID).
Patient height, weight and \gls{bmi} information were not available in this data set.
130 obesity-associated gene probes identified by \citet{Fuentes-Mattei2014} were taken from Supplementary Table 3 of their paper.

@@

The raw microarray gene expression data files from the \citet{Gatza2010a} study were downloaded from the \gls{geo} database (\gls{geo} accession ID: GSE1456, GSE\-1561, GSE2034, GSE3494, GSE4922 and GSE6596).
Only the Affymetrix HGU\-133A microarray samples were included in this project, as the other microarray data from the \citet{Creighton2012}, \citet{Fuentes-Mattei2014} and \citet{Print2016} studies were analysed using the Affymetrix HGU-133A platform.
Clinical data for the samples in \citet{Gatza2010a} study was not available, as these samples were a combination of many different datasets.
The Akt, \gls{bcat}, E2F1, \gls{egfr}, \gls{er}, \gls{her2}, \gls{ifna}, \gls{ifny}, Myc, p53, p63, \gls{pi3k}, \gls{pr}, Ras, \gls{stat3}, Src, \gls{tgfb}, and \gls{tnfa} pathway-associated genetic signatures were used by Gatza \textit{et al.}, and the gene probes of all the pathway-associated genetic signatures were obtained from Supplemental Table 10 of their paper.

@@

The raw microarray data were normalised using both \gls{rma} and \gls{mas} methods, each done separately on a copy of the data.
The reason that the \gls{mas} method was used as well as the \gls{rma} method was because the \gls{mas}-normalisation method was used in the \citet{Gatza2010a} study.
Though the normalisation methods were not specified in the \citet{Creighton2012} and \citet{Fuentes-Mattei2014} studies, to allow for data comparison between different datasets and accurate result validation of some of the studies, both \gls{rma} and \gls{mas} methods were considered for normalising the microarray data.

@@

In this project, two types of metagenes were created: obesity-associated and pathway-associated metagenes.
For obesity-associated metagenes, the direction of each metagene was checked with the patient obesity status and \gls{bmi} value from the data in which the metagene was derived from.
When the data set did not have \gls{bmi} or \gls{bmi} status information for the patients (for example the Gatza \textit{et al.} data set), a set of genes that were common in all obesity-associated genetic signatures were taken and the expression of these common genes were compared with the metagene scores.
The genes that are common to all of the obesity-associated genetic signatures will be related to the metagenes produced by these signatures, and therefore the expression patterns of these genes should be in line with the metagene scores.

Checking the direction of the pathway metagenes from \citet{Gatza2010a} study is not as simple as the obesity-associated metagenes, as the gene lists for each pathway were derived from a separate study and per-gene directionality information was not provided.
Due to this, it was not possible to validate whether the generated pathway metagene was in the ``correct'' direction.
As an alternative, the pathway metagenes were visually compared with the gene that represent the pathway using a heatmap, as the expression of this gene would likely be affected the most by the pathway signature.
For example, the \gls{er} pathway metagene was compared with the \textit{ESR1} gene expression, and if the metagene was not in the same direction as the \textit{ESR1} gene expression, the direction of the metagene was corrected (\cref{fig:methods/meta_dir}).
\cref{tab:metagene_direction} summarises the genes used to check the direction of each of the pathway metagenes.

@@

Since the \gls{nzbc} data was the only data set (other than the Creighton \textit{et al.} data set) that had the patient \gls{bmi} information, all of the linear models were constructed in the \gls{nzbc} data set.
Eleven linear models were created in the \gls{nzbc} to predict the obesity-associated metagene from the \citet{Creighton2012} study.
The first seven models created were combinations of the sample \gls{bmi}, \gls{bmi} status and a selection of pathway-associated metagene scores.
The selected pathways used in the linear model construction were \gls{bcat}, \gls{er}, \gls{ifna}, \gls{ifny}, Myc and \gls{pr}.
In deciding on the pathways to be used for the construction of the linear model, the correlation of the \gls{svd}- and transformation matrix-derived pathway metagene scores in different data sets were examined.

### 2017-10-12 thesis/misc/appendix.tex
Final revision completed

	Since all of these obesity-associated genetic signatures were derived based on the discrete values of sample \gls{bmi} staus, a genetic signature was identified using the sample \gls{bmi} value (continuous variable).
	Gene probe expressions were correlated with the sample \gls{bmi} values and was corrected for multiple hypothesis testing with \gls{fdr}.
	There were no significant gene that correlated with the sample \gls{bmi} (p \textless{} 0.05).

@@

To determine whether the results from all of the obesity metagene analyses were not showing any association with patient \gls{bmi} because of the \gls{svd}/transformation matrix approach used, the same approach was used with \gls{er} and \gls{pr} pathway-associated genetic signatures from the \citet{Gatza2010a} study.
Firstly, both the \gls{er} and \gls{pr} metagenes were generated in the \gls{rma} normalised CR data set.
The generated metagenes were compared with the gene expressions of the corresponding pathway-associated genetic signatures, and the association with the patient \gls{er} and \gls{pr} statuses were also examined.
As shown in \cref{fig:er_cr,fig:pr_cr}, both the \gls{er} and \gls{pr} metagenes in the CR data showed assocation with the gene expression pattern of the pathway, and also showed significant association with the patient \gls{er} and \gls{pr} statuses.

>>>>

	Since all of these obesity-associated genetic signatures were derived based on the discrete values of sample \gls{bmi} staus, a genetic signature was identified using the sample \gls{bmi} value (continuous variable).
	Gene probe expressions were correlated with the sample \gls{bmi} values and were corrected for multiple hypothesis testing with \gls{fdr}.
	There were no significant gene that correlated with the sample \gls{bmi} (p \textless{} 0.05).

@@

To determine whether the results from all of the obesity metagene analyses were not showing association with patient \gls{bmi} because of the \gls{svd}/transformation matrix approach used, the same approach was used with \gls{er} and \gls{pr} pathway-associated genetic signatures from the \citet{Gatza2010a} study.
Firstly, both the \gls{er} and \gls{pr} metagenes were generated in the \gls{rma} normalised CR data set.
The generated metagenes were compared with the gene expressions of the corresponding pathway-associated genetic signatures, and the association with the patient \gls{er} and \gls{pr} statuses were also examined.
As shown in \cref{fig:er_cr,fig:pr_cr}, both the \gls{er} and \gls{pr} metagenes in the CR data showed assocation with the gene expression pattern of the pathway, and also showed significant association with the patient \gls{er} and \gls{pr} statuses.
